Symbolic Weighted Language Models and Parsing over Infinite Alphabets

### 2021-05-11 main.tex
passe intro

%% theorem envs
\usepackage{theorem}
% \newtheorem{theorem}{Theorem} %[section]
% \newtheorem{definition}[theorem]{Definition}
% \newtheorem{lemma}[theorem]{Lemma}
% \newtheorem{corollary}[theorem]{Corollary}
% \newtheorem{proposition}[theorem]{Proposition}
% \newenvironment{proof}{\vspace{-2ex}{\it Proof. }}{\hspace*{\fill} $\Box$\smallskip }
% \theorembodyfont{\slshape}
% \newtheorem{example}[theorem]{Example}
% \newtheorem{remark}[theorem]{Remark}
{\theorembodyfont{\rmfamily} \theoremstyle{break} \newtheorem{algo}{Algorithm}}

@@

%the problem: quantitative parsing or symbolic parsing
%of parsing of words over infinite input alphabet.
Weighted parsing~\cite{Goodman99SemiringParsing,Nederhof03weightedParsing,MorbitzVogler19weighted-parsing} 
consists in analyzing a word in input \wrt a weighted language model, \eg a probabilistic context-free grammar (CFG).
In such models, a weight value is associated to each transition rule, % production rule
and the rule's weights can be combined with a associative product operator $\otimes$ into 
the weight of a derivation (or \emph{abstract syntax tree}, \ie the result of an analyze).
A second operator $\oplus$, associative and commutative, 
is moreover used to handle ambiguity of the model, 
by summing the weights of the possibly several (in general exponentially many) 
syntax trees associated to a given input word.
Intuitively, $\oplus$ aims at selecting, or ranking, the syntax trees.
The weight domain, equipped with these two operators is assumed, at minima, 
to form a \emph{semiring} $\Semiring$, %\cite{Eilenberg74automata}
such as a min-plus algebra of domain $\mathbb{R} \cup \{ +\infty\}$, 
where $\oplus$ is min and $\otimes$ is plus.
%by ranking 
%making the weight domain a semiring.
Some efficient specialized parsing algorithms~\cite{Huang05kbest} have been proposed in this context 
in order to compute the $n$ best syntax trees of a given input word without having to enumerate them all.
Generally based on dynamic programming, these algorithms rely on 
additional algebraic properties of~$\Semiring$.
-- see \eg~\cite{Huang05kbest} for some NLP applications.
%The extraction of $n$ best list is useful

The approaches to weighted parsing usually assume a finite input alphabet. %models and algorithms 
Considering infinite or large alphabets can however be of practical interest, 
for instance in the context of processing strings in UTF16 encodings, 
for vulnerability detection in Web-applications 
(see \eg~\cite{dAntoni21CACM}), 
for the representation of execution traces in model checking, 
or when dealing with sequences of timestamped events
-- we describe briefly at the end of the paper a parsing based approach to 
of automated music transcription, that is the conversion of a symbolic 
representation of a music performance into a structured music score.
%\cite{MCM19} 
%generalizes weighted parsing: 
%finding the best derivation for a weighted grammar.

We present here some weighted language models
%model of weighted CFG 
computing on words over infinite input alphabets %sets of terminal symbol
- but with finite sets of states and transitions rules. 
% non-terminals and production rules).
In their transition rules, input symbols appear as variables %(parameters)
and the weight associated to a transition rule is a function of these variables.
% 
This approach is close to the case of 
Symbolic Automata (SA)~\cite{dAntoniVeanes17CAV,dAntoni21CACM}, 
except that the domain for weight values is not restricted to be Boolean, 
like for the guards in the rules of SA, 
but can be an arbitrary commutative semiring (assuming some restrictions).
%
The models defined here are finite automata (\SWA),
transducers (\SWT) and pushdown automata
with a visibly restriction~\cite{AlurMadhusudan09nested} (\SWVPA).
%wrt an edit distance
The latter model of automata computes on \emph{nested words}~\cite{AlurMadhusudan09nested}, 
a structured form of words parenthesized with markup symbols, 
corresponding to a linearization of trees.
In expressiveness, they correspond to a strict restriction
of weighted CFG. 
%and can be used in a general approach for parsing over infinite input alphabets.
Let $A$ be a \SWVPA, associating $A(w) \in \Semiring$ 
to a given a nested word $w$ (representing a parse tree),
and let a \SWT compute a distance $d$, in $\Semiring$, 
between 2 strings over respectively an infinite input alphabet and the 
same (infinite) alphabet of $A$.
Then, the problem of Symbolic Weighted Parsing is, 
given an input string $s$, to find a nested word $w$ minimizing 
(according to the ranking defined by $\oplus$)
the distance $d(s, w) \otimes A(w)$ between $s$ and $A$, 
as defined in~\cite{Mohri03EDWA}.
%following

@@

\subsection{Definitions} \label{sec:SWTdef}\label{sec:SWAdef}
%(SWT)
\begin{definition}
\label{def:transducer} \label{def:SWT}
A \emph{symbolic-weighted transducer} (\SWT)
over the input and output alphabets~$\Sigma$ and $\Delta$ 
and the semiring $\Semiring$ is a tuple
$T = \< Q, \init, \bar{\wei}, \final >$,
where $Q$ is a finite set of states, 
$\mathsf{in} : Q \to \Semiring$, 
respectively $\mathsf{out} : Q \to \Semiring,$
are functions defining the weight for entering, 
respectively leaving, a state, 
and $\bar{\wei}$ is a tuplet made of the 4 of transition functions 
$\wei_\epsilon$, $\wei_\Sigma$, $\wei_\Delta$, and $\wei_{\Sigma, \Delta}$
from $Q \times Q$ into respectively~$\Semiring$, %$\Phi_\epsilon$, 
$\Phi_\Sigma$, $\Phi_\Delta$, and $\Phi_{\Sigma, \Delta}$.
\end{definition}
%
We summerize the above 4-uplet of transition functions into a unique 
function~$\wei$ from
$Q \times (\Sigma \cup \{ \epsilon \}) \times (\Delta \cup \{ \epsilon \}) \times Q$
into $\Semiring$, %also called $\wei$ for simplicity, 
such that, for all $q, q' \in Q$, $a \in \Sigma$, and $b \in \Delta$, 
%and with 
%$\< \phi_\epsilon, \phi_\Sigma, \phi_\Delta, \phi_{\Sigma, \Delta}> = \wei(q, q')$, 
\[
\begin{array}{rcll}
\wei(q, \epsilon, \epsilon, q') & = &  \wei_\epsilon(q, q'),\\ %\phi_\epsilon\\
\wei(q, a, \epsilon, q') & = & \phi_\Sigma(a) & %\wei_\Sigma(q, q')(a)
\quad\mathrm{where~} \phi_\Sigma = \wei_\Sigma(q, q'),\\
\wei(q, \epsilon, b, q') & = & \phi_\Delta(b) & 
\quad\mathrm{where~} \phi_\Delta = \wei_\Delta(q, q'),\\
\wei(q, a, b, q') & = & \phi_{\Sigma, \Delta}(a, b) & 
\quad\mathrm{where~} \phi_{\Sigma, \Delta} = \wei_{\Sigma, \Delta}(q, q').\\
\end{array}      
\]
%More precisely, $Q \times Q$, 
%resp. $Q \times \Sigma \times Q$,
%$Q \times \Delta \times Q$,
%$Q \times \Sigma \times \Delta \times Q$,
%into resp. $\Phi_0$, $\Phi_\Sigma$, $\Phi_\Delta$ $\Phi_{\Sigma, \Delta}$.

@@

\begin{example}
(simpler) poinwise distance between two sequences of timestamped events **
\end{example}

@@

Let $\Sigma$ be a countable input alphabet 
and $\Omega = \Omegai \uplus \Omegac \uplus \Omegar$ an output countable alphabet,
let $\< \Semiring, \oplus, \zero, \otimes, \one>$ be a commutative, bounded, complete 
and total semiring 
and let  $(\Phi_\epsilon, \Phic, \Phir, \Phicr)$ be a label theory over $\Semiring$, 
assumed computable and $k$-convex for some fixed $k$.
%
\noindent
Assuming given in input:
\begin{itemize}
\item a \SWT $T$ over $\Sigma$,  $\Omega$, and $\Semiring$,
defining a measure %between words 
$T: \Sigma^* \times \Omega^* \to \Semiring$,
\item a \SWVPA $A$ over $\Omega$, and $\Semiring$, defining a series of nested words
      $A : \Omega^* \to \Semiring$,
\item an input word $s \in \Sigma^*$,
\end{itemize}
the problem of \emph{symbolic weighted parsing} is to find 
a nested word $t \in \Omega$ minimizing
\( T(s, t) \otimes A(t)\)
\wrt $\leq_\oplus$, 
\ie such that %Hence, it is the problem of finding 
\( T(s, t) \otimes A(t) = 
  \displaystyle\bigoplus_{t' \in \Omega^*} T(s, t') \otimes A(t') \).

\noindent 
Therefore, it is the problem of optimizing a measure 
called the \emph{edit-distance between $s$ and $A$} in~\cite{Mohri03EDWA}.
%
The input language can also be expressed as a \SWTA, or, 
as a particular case, as a weighted context-free grammar, 
converted in turn into a \SWVPA following Lemma~\ref{lem:SWTA}.
%
In the case of finite alphabets, 
the problem of searching, in a WTA language, 
for the best parse tree for a given,  
sometimes referred as~\emph{weighted parsing}
(see \cite{Goodman99semiring,MorbitzVogler19weighted-parsing} **more general problems**)
is a particular case of SW parsing.
%
Indeed, it corresponds to the case where $T$
accepts only the pairs $\<s, t>$ such that 
$s$ is the projection of $t$ on $\Omegai$. 
This can be done with a single state $q$ and 
with transition rules of the form:
\begin{description}
\item[] $\wei(q, \epsilon, a, q) = \one$ for all $a \in \Omegac \cup \Omegar$,
\item[] $\wei(q, a, a, q) = \one$ for all $a \in \Omegai$,
\item[] $\wei(q, a, b, q) = \zero$ for all $a, b \in \Omegai$, $a \neq b$.
\end{description}

>>>>

%% theorem environments
\usepackage{theorem}
% \newtheorem{theorem}{Theorem} %[section]
% \newtheorem{definition}[theorem]{Definition}
% \newtheorem{lemma}[theorem]{Lemma}
% \newtheorem{corollary}[theorem]{Corollary}
% \newtheorem{proposition}[theorem]{Proposition}
% \newenvironment{proof}{\vspace{-2ex}{\it Proof. }}{\hspace*{\fill} $\Box$\smallskip }
% \theorembodyfont{\slshape}
% \newtheorem{example}[theorem]{Example}
% \newtheorem{remark}[theorem]{Remark}
{\theorembodyfont{\rmfamily} \theoremstyle{break} \newtheorem{algo}{Algorithm}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% intro
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction} \label{sec:intro}
Parsing %variant of membership problem for formal languages
%- given a language model and an input word $s$, compute a derivation of the model that yields $s$.
is the problem %process 
of structuring an input linear representation %typically
(a finite word over an alphabet) according to a language model (a formal grammar).
% natural language, programming language, 
%
Most of the context-free parsing approaches~\cite{GruneJacobs08parsing}  
assume a finite and reasonably small input alphabet. %models and algorithms 
Such a restriction makes perfectly sense in the context of NLP tasks 
like constituency parsing or of programming languages compilers or interpreters.
Considering large or infinite alphabets can however be of practical interest
in other cases.
%
For instance, when dealing with large characters encodings such as UTF-16, % processing strings in 
\eg for vulnerability detection in Web-applications~\cite{dAntoni21CACM}, 
%
for the analyse of data streams or serialization of structured documents~\cite{Segoufin06csl}, 
or processing of timed execution traces.
Regarding the latter case, we describe briefly at the end of the paper 
a parse-based approach to automated music transcription,
where a symbolic music performance (sequence of timed musical events) % timestamped
% a symbolic representation of a music performance 
is converted into a structured score in Common Western music notation~\cite{foscarin:hal-01988990}.
%generalizes weighted parsing: 
%finding the best derivation for a weighted grammar.

Various extensions of automata have been studied for handling infinite alphabets.
%words carrying data values in an infinite domain (e.g. integers) e.g. data processing 
To cite a few, some memory extensions allow restricted storage and comparison of input symbols, 
%and correspond with logics, 
(see~\cite{Segoufin06csl} for a survey)
like pebbles for marking positions~\cite{NevenSchwentickVianu04FSMinfinite}, 
registers~\cite{KaminskiFrancez94}, 
computations in 2 steps~\cite{Bojanczyk11FO2}... % data words automata.
%
%for the and verification of infinite state systems 
%(model checkers: alphabet = long bit-vectors)
%...For the representation of  in model checking, verification and 
similar to representation of automata in model checkers, 
following a similar idea, 
Symbolic automata, 
transitions guarded by predicates over alphabet theories.

another quantitative issue in parsing is non-determinism : 
with an ambiguous grammar, there may exists several derivations yielding one input word.
%the problem: quantitative parsing or symbolic parsing
%of parsing of words over infinite input alphabet.
selection by ranking derivation according to an associated weight value...
Weighted parsing~\cite{Goodman99SemiringParsing,Nederhof03weightedParsing,MorbitzVogler19weighted-parsing} 
consists in analysing a word in input \wrt a weighted language model, \eg a probabilistic context-free grammar (CFG).
In such models, a weight value is associated to each transition rule, % production rule
and the rule's weights can be combined with a associative product operator $\otimes$ into 
the weight of a derivation (or \emph{abstract syntax tree}, \ie the result of an analyze).
A second operator $\oplus$, associative and commutative, 
is moreover used to handle ambiguity of the model, 
by summing the weights of the possibly several (in general exponentially many) 
syntax trees associated to a given input word.
Intuitively, $\oplus$ aims at selecting, or ranking, the syntax trees.
The weight domain, equipped with these two operators is assumed, at minima, 
to form a \emph{semiring} $\Semiring$, %\cite{Eilenberg74automata}
such as a min-plus algebra of domain $\mathbb{R} \cup \{ +\infty\}$, 
where $\oplus$ is min and $\otimes$ is plus.
%by ranking 
%making the weight domain a semiring.
Some efficient specialized parsing algorithms~\cite{Huang05kbest} have been proposed in this context 
in order to compute the $n$ best syntax trees of a given input word without having to enumerate them all.
Generally based on dynamic programming, these algorithms rely on 
additional algebraic properties of~$\Semiring$.
-- see \eg~\cite{Huang05kbest} for some NLP applications.
%The extraction of $n$ best list is useful

We present here some weighted language models
%model of weighted CFG 
computing on words over infinite input alphabets %sets of terminal symbol
- but with finite sets of states and transitions rules. 
% non-terminals and production rules).
In their transition rules, input symbols appear as variables %(parameters)
and the weight associated to a transition rule is a function of these variables.
% 
This approach is close to the case of 
Symbolic Automata (SA)~\cite{dAntoniVeanes17CAV,dAntoni21CACM}, 
except that the domain for weight values is not restricted to be Boolean, 
like for the guards in the rules of SA, 
but can be an arbitrary commutative semiring (assuming some restrictions).
%
The models defined here are finite automata (\SWA),
transducers (\SWT) and pushdown automata
with a visibly restriction~\cite{AlurMadhusudan09nested} (\SWVPA).
%\wrt an edit distance
The latter model of automata computes on \emph{nested words}~\cite{AlurMadhusudan09nested}, 
a structured form of words parenthesized with markup symbols, 
corresponding to a linearization of trees.
In expressiveness, they correspond to a strict restriction
of weighted CFG. 
%and can be used in a general approach for parsing over infinite input alphabets.
Let $A$ be a \SWVPA, associating $A(w) \in \Semiring$ 
to a given a nested word $w$ (representing a parse tree),
and let a \SWT compute a distance $d$, in $\Semiring$, 
between 2 strings over respectively an infinite input alphabet and the 
same (infinite) alphabet of $A$.
Then, the problem of Symbolic Weighted Parsing is, 
given an input string $s$, to find a nested word $w$ minimizing 
(according to the ranking defined by $\oplus$)
the distance $d(s, w) \otimes A(w)$ between $s$ and $A$, 
as defined in~\cite{Mohri03EDWA}.
%following

@@

\subsection{Definitions} \label{sec:SWTdef}\label{sec:SWAdef}
%(SWT)
\begin{definition}
\label{def:transducer} \label{def:SWT}
A \emph{symbolic-weighted transducer} (\SWT)
over the input and output alphabets~$\Sigma$ and $\Delta$ 
and the semiring $\Semiring$ is a tuple
$T = \< Q, \init, \bar{\wei}, \final >$,
where $Q$ is a finite set of states, 
$\mathsf{in} : Q \to \Semiring$, 
respectively $\mathsf{out} : Q \to \Semiring,$
are functions defining the weight for entering, 
respectively leaving, a state, 
and $\bar{\wei}$ is a tuplet made of the 4 of transition functions 
$\wei_\epsilon$, $\wei_\Sigma$, $\wei_\Delta$, and $\wei_{\Sigma, \Delta}$
from $Q \times Q$ into respectively~$\Semiring$, %$\Phi_\epsilon$, 
$\Phi_\Sigma$, $\Phi_\Delta$, and $\Phi_{\Sigma, \Delta}$.
\end{definition}
%
We summarize the above 4-uplet of transition functions into a unique 
function~$\wei$ from
$Q \times (\Sigma \cup \{ \epsilon \}) \times (\Delta \cup \{ \epsilon \}) \times Q$
into $\Semiring$, %also called $\wei$ for simplicity, 
such that, for all $q, q' \in Q$, $a \in \Sigma$, and $b \in \Delta$, 
%and with 
%$\< \phi_\epsilon, \phi_\Sigma, \phi_\Delta, \phi_{\Sigma, \Delta}> = \wei(q, q')$, 
\[
\begin{array}{rcll}
\wei(q, \epsilon, \epsilon, q') & = &  \wei_\epsilon(q, q'),\\ %\phi_\epsilon\\
\wei(q, a, \epsilon, q') & = & \phi_\Sigma(a) & %\wei_\Sigma(q, q')(a)
\quad\mathrm{where~} \phi_\Sigma = \wei_\Sigma(q, q'),\\
\wei(q, \epsilon, b, q') & = & \phi_\Delta(b) & 
\quad\mathrm{where~} \phi_\Delta = \wei_\Delta(q, q'),\\
\wei(q, a, b, q') & = & \phi_{\Sigma, \Delta}(a, b) & 
\quad\mathrm{where~} \phi_{\Sigma, \Delta} = \wei_{\Sigma, \Delta}(q, q').\\
\end{array}      
\]
%More precisely, $Q \times Q$, 
%resp. $Q \times \Sigma \times Q$,
%$Q \times \Delta \times Q$,
%$Q \times \Sigma \times \Delta \times Q$,
%into resp. $\Phi_0$, $\Phi_\Sigma$, $\Phi_\Delta$ $\Phi_{\Sigma, \Delta}$.

@@

\begin{example}
(simpler) pointwise distance between two sequences of timestamped events **
\end{example}

@@

Let $\Sigma$ be a countable input alphabet 
and $\Omega = \Omegai \uplus \Omegac \uplus \Omegar$ an output countable alphabet,
let $\< \Semiring, \oplus, \zero, \otimes, \one>$ be a commutative, bounded, complete 
and total semiring 
and let  $(\Phi_\epsilon, \Phic, \Phir, \Phicr)$ be a label theory over $\Semiring$, 
assumed computable and $k$-convex for some fixed $k$.

\noindent
Assuming given in input:
\begin{itemize}
\item a \SWT $T$ over $\Sigma$,  $\Omega$, and $\Semiring$,
defining a measure %between words 
$T: \Sigma^* \times \Omega^* \to \Semiring$,
\item a \SWVPA $A$ over $\Omega$, and $\Semiring$, defining a series of nested words
      $A : \Omega^* \to \Semiring$,
\item an input word $s \in \Sigma^*$,
\end{itemize}
the problem of \emph{symbolic weighted parsing} is to find 
a nested word $t \in \Omega$ minimizing
\( T(s, t) \otimes A(t)\)
\wrt $\leq_\oplus$, 
\ie such that %Hence, it is the problem of finding 
\( T(s, t) \otimes A(t) = 
  \displaystyle\bigoplus_{t' \in \Omega^*} T(s, t') \otimes A(t') \).

\noindent 
Therefore, it is the problem of optimizing a measure 
called the \emph{edit-distance between $s$ and $A$} in~\cite{Mohri03EDWA}.
%
The input language can also be expressed as a \SWTA, or, 
as a particular case, as a weighted context-free grammar, 
converted in turn into a \SWVPA following Lemma~\ref{lem:SWTA}.
%
In the case of finite alphabets, 
the problem of searching, in a WTA language, 
for the best parse tree for a given,  
sometimes referred as~\emph{weighted parsing}
(see \cite{Goodman99SemiringParsing,MorbitzVogler19weighted-parsing} **more general problems**)
is a particular case of SW parsing.
%
Indeed, it corresponds to the case where $T$
accepts only the pairs $\<s, t>$ such that 
$s$ is the projection of $t$ on $\Omegai$. 
This can be done with a single state $q$ and 
with transition rules of the form:
\begin{description}
\item[] $\wei(q, \epsilon, a, q) = \one$ for all $a \in \Omegac \cup \Omegar$,
\item[] $\wei(q, a, a, q) = \one$ for all $a \in \Omegai$,
\item[] $\wei(q, a, b, q) = \zero$ for all $a, b \in \Omegai$, $a \neq b$.
\end{description}
